---
author: "YOUR NAME HERE"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---
# Lab 1 - Flow analysis and hydrograph separation (20 pts)

## Intro

In this lab we will analyze stream flow (Q) and precipitation (P) data from Tenderfoot Creek Experimental Forest (TCEF). TCEF is located in central Montana, north of White Sulphur Springs. [See here](https://www.fs.usda.gov/research/rmrs/forestsandranges/locations/tcef){target="_blank"} for information about TCEF. You will do some data analysis on flows, calculate annual runoff ratios, and perform a hydrograph separation. 

## Reading for this lab

Ladson, A. R., R. Brown, B. Neal and R. Nathan (2013) A standard approach to baseflow separation using the Lyne and Hollick filter. Australian Journal of Water Resources 17(1): 173-18

+ [Ladson et al., 2013](https://www.researchgate.net/publication/269755620_A_standard_approach_to_baseflow_separation_using_the_Lyne_and_Hollick_filter){target="_blank"}

Lynne, V., Hollick, M. (1979) Stochastic time-variable rainfall-runoff modelling. In: pp. 89-93 Institute of Engineers Australia National Conference. Perth.

+ [Lyne and Hollick, 1979](https://www.dropbox.com/scl/fi/vzuy13znsou4m149e0xkx/Stochastic_Time-Variable_Rainfall-Runoff.pdf?rlkey=rztj4c1nd3xhy765eac7mav5j&dl=0){target="_blank"}

## Repo

Follow this [link](https://github.com/tpcovino/01_hydrograph_sep){target="_blank"} to download everything you need for this unit. When you get to GitHub click on "Code" (green button) and select "download zip". You will then save this to a local folder where you should do all of your work for this class. You will work through the "_blank.Rmd" or "_partial.Rmd". Always be sure to read the README.md files in the GitHub repo. Sometimes they are useful, sometimes they aren't, but always have a look. 

As I mentioned above you will work through the "_blank.Rmd" or "_partial.Rmd". If you want to learn how to code in R, I encourage you to work through the blank version as much as possible. Also, if you don't have much R background this lab might seem kind of challenging. But don't worry. I'm challenging you right now, but it will get easier, and you can always reach out to us with questions and we can do video chats if a live demonstration is needed. So don't get frustrated if this seems tough right now. Soon you will be rattling off code with ease. Conversely, if you are an experienced coder and have ideas for how to do this in ways other than what I've shown here, please share code with your colleagues and help them develop their coding skills! 

OK. Once you have this folder saved where you would like it, open RStudio and navigate to the folder. Next, open the project (".Rproj"). Doing so will set the folder as the working directory, make your life easier, and make everything generally work. The use of projects is highly recommended and is the practice we will follow in this class. See [here](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) for an overview of projects and why you should use them from Jenny Bryan. 

If you are new to R, or need a refresher, please read [Chapters 1, 2, & 3](https://r4ds.had.co.nz/introduction.html){target="_blank"} 1-Welcome, 2-Introduction, & 3-Data visualization in R for Data Science (RDS).

In this unit we want to start familiarizing ourselves with R by visualizing some hydrological data.

## Knitr

This is knitr settings. Knitr is a package that will turn this Rmd to an html. 
```{r}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, results = FALSE)

```

## Packages

Here, we will load some packages. If you have never installed these you will need to install then load. This would look like:

install.packages("tidyverse")
library(tidyverse)

The tidyverse is a collection of packages. [See here](https://www.tidyverse.org/packages/) for an overview of the Tidyverse. 

The second package below (lubridate) actually loads with the Tidyverse, but I wanted to explicitly bring your attention to lubridate. Lubridate is a package for working with dates. Dates can be frustrating in R, and lubridate helps limit that frustration. You will often see that when you load data into R, the dates will be characters, not dates. We will deal with this in this lab, and it is something you should always check when you load data. 

```{r setup, include = FALSE}
library(tidyverse)
library(lubridate) # loads with tidyverse, don't actually need this line
library(hydrostats) # used for hydrograph separation.
library(plotly) # used as an extension to ggplot that makes plots interactive
```

## Tenderfoot Analysis - Part A

In this lab you will use precipitation and runoff data from Tenderfoot Creek (TCEF) in central Montana. TCEF has a snow dominated hydrologic regime and has one big pulse of water from snowmelt. The dataset ("PQdata.csv") contains P and Q data from three watersheds in TCEF. You are asked to do some basic data analysis and a hydrograph separation at the end. Follow the instructions for your analysis. Additionally, answer the questions that are interspersed.

### Load the data with `read_csv()`.

```{r}
rm(list = ls()) # this clears out the Environment ("fresh start!")
df <- BLANK("PQdata.csv") # read in data
view(df)
str(df)
```

Look at the df. "LSC", "BUB", and "UTC" are watersheds in the TCEF. "P" is precipitation (snowmelt and rain), "Q" is streamflow. The date column is actually a character and not formatted as a date. We first need to make sure that the "Date" column has the proper format. Use the `lubridate` package and the appropriate function to modify the "Date" column with `mutate()`.

```{r}
df <- df %>% 
  mutate(Date = BLANK(Date))
str(df)
view(df)
```

In hydrology, we use the water year (WY) instead of the calendar year. As a reminder, the WY begins October 1 and ends September 30 the following year. So, 2014-10-01 through 2015-09-30 is the 2015 WY. Add an additional column to the large df that contains just a number for the water year. Use `if_else()` for this. It works like this: if_else(If some condition is met, do this, otherwise do this). 

Look at the following example to understand the `if_else()` function.
```{r}
testvector <- c(1,2,5,8,3,6,9,0,3,1,5,7,6,9,1,0)
new_testvector <- if_else(testvector >= 5, 10, 0)
new_testvector
```

**1) (1 pt) What happened here?**  
ANSWER: 

Now use `if_else()` to create a water year column. The easiest way is to look at the individual months and put the months October, November, and December into the next year. So, our instruction is: if_else(if the month is greater than 9, add 1 to the year, otherwise just leave the year). You can extract the month from the "Date" column with `month(Date)` and the year with `year(Date)`.

```{r}
df <- df %>% 
  mutate(WY = BLANK(month(BLANK) > 9, year(Date) + 1, year(Date))) %>% # this adds a column that contains the water year
  relocate(WY, .after = Date)
view(df)
```

**2) (2 pts) Why is it important (especially in the western US) to use the water year instead of the calendar year for hydrologic analysis? Some places in the east use a May-April water year instead. Why do you think that is?**   
ANSWER:

Let's plot some raw data. How would you plot Q for the three watersheds to do some exploratory data analysis (EDA)? **Note:** EDA should always be your first step after loading some data. 

I, personally, use ggplot for this. Plotting with ggplot looks something like:

ggplot(data = df, aes(x = x, y = y)) + 
geom_point()

And other commands. See [here](https://ggplot2.tidyverse.org/) for documentation on ggplot. 

```{r}
df %>% 
  ggplot(aes(x = BLANK)) +
  geom_line(aes(y = LSC_Q, color = "LSC")) + 
  geom_line(aes(y = BLANK, color = "BUB")) + 
  geom_line(aes(y = UTC_Q, color = "UTC")) +
  labs(color = "Watershed", y = "Q (mm/hr)", x = "")
```

Because we have made a year column using "mutate" we can filter for only the 2004 WY. That would look something like:

df_2004 <- df %>% 
filter(year == 2004)

First, make a data frame called df_2004 with only data from WY 2004. Have a look at the data. 

```{r}


```

OK - so the data for the 2004 WY is missing, so let's remove it from the time series. We'll also remove the 2011 data because of potential measurement errors in one of the watersheds. `drop_na()` removes all rows that contain NAs (that will get rid of 2004). In this case `drop_na()` does not require any other inputs in the parentheses. Use `filter()` to get rid of the 2011 water year. `filter()` works by selecting rows to keep. So, to get rid of one particular year, you effectively have to select all other years. However, this can be simplified by saying "keep all water years that are NOT 2011". The operator for "not equal to" in R is `!=`. If you wanted to choose specific years to keep, you could do this with the `==` operator ("exactly equal to"). Note: You could also use `filter()` to remove the 2004 WY, but with `drop_na()` you learned another function.

```{r}
df_BLANK<- df %>% 
BLANK(year == BLANK)
```

**3) (2 pt) What would the `filter()` code look like if you wanted to choose only WYs 2000 and 2008? What is the difference in R between NA and NaN?**    
ANSWER: 

### Make tidy data
The next step is to bring the data to the tidyr long form. Also rename the df to "PQ". The next chunk appears complicated, but is pretty logical if you follow the operations.

[Here](https://r4ds.hadley.nz/data-tidy) is a chapter on data tidying. 

```{r}
PQ <- df %>% 
  # make a long df by excluding Date and WY from the restructuring
  pivot_longer(c(-Date, -WY), names_to =  "Key", values_to =  "Value") %>%
  # separate the Key column into the watershed name and water cycle component
  separate(Key, into = c("Watershed", "Flux"), sep =  "_") %>% 
  # now turn the Flux column back into two separate columns
  pivot_wider(names_from =  "Flux", values_from = "Value")
  
```

### Daily Q plot

Plot the daily Q time series for the three watersheds with "Date" on the x axis and "Q" on the y axis. Use `color = ` to distinguish between the three watersheds. Also label the y-axis appropriately as "Q (mm/day)". Feel free to to change other default settings to make the graph look nicer, too. Use `scale_x_date()` to place the ticks and labels every two years. We haven't looked at this yet, so you might have to do some googling.

```{r}

q_plot <- ggplot() +
  geom_line(data = PQ, aes(x = Date, y = Q, color = BLANK)) +
  labs(y = "Q (mm/day)") +
  BLANK(date_breaks = "2 years", date_labels = "%Y") +
  scale_color_viridis_d()

ggplotly(BLANK)

```

### Annual stats

Calculate annual total P, Q, ET, and RR (runoff ratios) for all three watersheds. The grouping needs to be done both on the Watershed AND WY. Use `summarise()` to calculate the values. Also change the watershed names from characters to factors using as_factor. See [here](https://forcats.tidyverse.org/reference/as_factor.html). 

In addition to summarise this operation can also be accomplished with reframe. See [here](https://dplyr.tidyverse.org/reference/reframe.html) for information on reframe.

```{r}

PQ_annual <- PQ %>%
  group_by(Watershed, WY) %>% 
  BLANK(Qsum = sum(Q), Psum = sum(BLANK), RR = Qsum/Psum) %>% 
  ungroup() %>% 
  mutate(Watershed = as_factor(Watershed))

```

### Annual RR plot

Plot the annual runoff ratios for the three watersheds with the confidence bands in the same color as the line.
```{r}

ggplot(data = PQ_annual, aes(x = WY, y = RR, color = Watershed, fill = Watershed)) +
  theme_bw(base_size = 15) +
  geom_point() +
  BLANK(method = "loess") +
  scale_color_viridis_d() +
  scale_fill_viridis_d()

```

**4) (1 pt) What does `color` do, what does `fill` do? What is `method = "loess"?**  
ANSWER


**5) ((2 pt) Describe the behavior of the three lines. What do you think affects a watershed's annual runoff ratio? Feel free to generate more plots to guide your answer.**)  
ANSWER

**6) (2 pts) Let's assume the watersheds are similar to each other (geology, soils, vegetation), but they differ in slope. Rank the watersheds in steepness and explain why you ranked them this way**  
ANSWER:


## Hydrograph separation - Part B
### Full timeseries

We will use an existing R package to perform a hydrograph separation. First we need to get the data in shape. Let's remove the precipitation data and the water year column using `select()`, we only want the Date and Q for all three watersheds. Then, with `pivot_wider()`, turn the df from long into wide format by spreading out the Q data for each watershed. We want the df to have 4 columns: Date, BUB, LSC, UTC, with the last three columns being the runoff for the respective watershed. 

Note BUB, LSC, and UTC are watersheds in TCEF. These are Bubbling (BUB), Lower Stringer Creek (LSC), and Upper Tenderfoot Creek (UTC).

`pivot_wider()` has similar syntax to `pivot_longer()`, but here the important parts are `names_from =` and `values_from =`. In this case, the names come from the Watershed column (which is now a factor), the values come from the Q column.

Use pivot_wider() to make a df called Qdat that has date and Q for each watershed in the so-called "wide" format. See [here](https://r4ds.had.co.nz/tidy-data.html) and [here](https://anvil.works/blog/tidy-data) for information on wide vs long data formats. Note that the second link is a Python article. So we aren't using those packages and we won't in this class. But point being "long" vs. "wide" is relevant in all the programming languages, including Python and R.   

```{r}
Qdat <- PQ %>% 
  select(-P, -WY) %>% 
  BLANK(names_from = Watershed, values_from = Q)
view(Qdat)
```

The function for hydrograph separation spits out annual data on a calendar year basis. We don't want that. Hence, we calculate everything on a daily basis and then aggregate manually to water years. Let's first do the hydrograph separation for the entire timeseries to get the overall means. The `baseflows()` function uses the Lyne & Hollick, 1979, filter. 

+ [Lyne and Hollick, 1979](https://www.dropbox.com/scl/fi/vzuy13znsou4m149e0xkx/Stochastic_Time-Variable_Rainfall-Runoff.pdf?rlkey=rztj4c1nd3xhy765eac7mav5j&dl=0){target="_blank"}

Most of the code is provided, but try to understand what it does. I put the calculation of the hydrograph separation into a for loop.
```{r}
HS_summary <- NULL # preallocate the hydrograph separation df
for (i in 1:3) { # for all three watersheds
  # the next three lines take the Q for one watershed and write it into a new df. The `baseflows()` function needs two columns: Date and Q.
  Qbf <- Qdat %>% 
    select(Date, names(Qdat[,(i + 1)])) %>% # pick the date and Q columns for a watershed
    rename(Q = names(Qdat[(i + 1)])) # rename the column with runoff "Q"
  
  bf_filter <- 0.975 # set filter parameter
  temp <- baseflows(Qbf, bf_filter, n.reflected = 30, ts = "mean") # actual baseflow separation
  # the next lines stitch the results from all runs together
  temp1 <- bind_cols(temp, # take the current bseflow separation
                     as_tibble( # convert to tibble
                       rep( # replicate
                         names( # the watershed name
                           Qdat[(i + 1)]), # grab the current watershed name
                         dim(temp)[1]))) # as often as there are rows in the baseflow separation
  HS_summary <- bind_rows(HS_summary, temp1) # put all bf seps together
}

# make a clean looking df with all results from the baseflow separations
HS_summary <- HS_summary %>% 
  mutate(mean.qfi = 1 - mean.bfi) %>% # this add a value for quickflow (stormflow) percentage (1 - baseflow)
  rename(Watershed = value)


# create long form for plotting with only quickflow and baseflow and the watershed names
HS_long <- HS_summary %>%
  select(mean.bfi, mean.qfi, Watershed) %>% # mean.bfi, mean.qfi, Watershed
  pivot_longer(names_to = "key", values_to = "value", -Watershed) # make the long form. Ends up with three columns.

# plot column/bar graph
bf_plot <- ggplot() +
  geom_col(data = HS_long, aes(x = Watershed, y = value, fill = key), width = 0.7) + # plot percentage of bfi/qfi for the three watersheds
  scale_fill_discrete(labels = c("Baseflow", "Stormflow")) + # change the legend labels to represent "Baseflow" and "Stormflow"
  labs(x = "Watershed", y = "Fraction", fill = "Flow") # label the axes and legend
  

ggplotly(bf_plot)

```


**7) (2 pts) Look at the results of the hydrograph separation. Do the percentages of the individual watersheds make sense relative to each other? Think of the answer to the previous question.**   
ANSWER: 


**8) (1 pt) What happens if you modify the filter parameter?**  
ANSWER:


#### Annual hydrograph separations
The example above calculates the hydrograph separation for the entire 16 year time series. We can do some more analysis on the annual data. For example, we can check if the annual values of BFI are correlated with the annual runoff ratios. First, we need to get the BFI for the watersheds for every year. `ts` needs to be set to "daily".

Alter the hydrograph separation code from above to now compute the separation at the daily timestep (ts). And then follow code similar to the RR code to calculate annual mean BFI from the daily data. 

```{r}

HS_daily <- NULL # preallocate 
for (i in 1:3) {
  Qbf <- Qdat %>% 
    select(Date, names(Qdat[,(i + 1)])) %>% 
    rename(Q = names(Qdat[(i + 1)]))
  
  bf_filter <- 0.975 # set filter parameter
  temp <- baseflows(Qbf, bf_filter, n.reflected = 30, ts = BLANK) # actual baseflow separation
  temp1 <- bind_cols(temp, as.tibble(rep(names(Qdat[(i + 1)]), dim(temp)[1]))) # append the watershed name to the bs separation result
  HS_daily <- bind_rows(HS_daily, temp1) # put all bf seps together
}

# add a column with the water year
HS_daily <- HS_daily %>% 
  mutate(WY = if_else(month(Date) > 9, year(Date) + 1, year(Date))) %>% # this adds a column that contains the water year
  rename(Watershed = value)

# group the daily data by water year and watershed and then calculate the mean bfi for the year (call the column "mean.bfi").
HS_annual <- HS_daily %>% 
  group_by(WY, Watershed) %>% 
  # this just calculates the mean of the daily baseflow values
  # summarise(mean.bfi = mean(bfi))
  # comment out the previous line and uncomment for total baseflow values based on actual runoff
  summarise(sum.bf = sum(bf), sum.Q = sum(Q)) %>% # calculates baseflow totals and runoff totals
  mutate(mean.bfi = sum.bf/sum.Q) # real bfi is total baseflow sum / total runoff

```

Now we have two dfs with 42 observations for the watersheds, one holding the BFI means (HS_annual), the other the values for P, Q, ET, and RR (PQ_annual). Now we need to put those two together. We can use `right_join()` from `dplyr` to combine the two dfs into a new df called PQ_all. Once you have created PQ_all, make a plot with annual BFI on the y, and annual RR on the x. Color by watershed. And add a trend line with 95 confidence intervals by using stat_smooth. 
```{r}
PQ_all <- BLANK(HS_annual, PQ_annual) # combine the dfs

# plot the data
bfi_annual_plot <- ggplot(data = PQ_all, aes(x = RR, y = mean.bfi, color = Watershed, fill = Watershed)) + # color determines the line color, fill the confidence band color
  theme_bw(base_size = 15) +
  geom_point(shape=16, size=4, alpha=0.5) + # plot as points
  stat_smooth(BLANK = "lm", alpha = 0.1) + # plot linear regression with confidence bands
  labs(x = "Annual Runoff Ratio", y = "Annual Baseflow Index") +
  scale_fill_viridis_d() +
  scale_color_viridis_d()

ggplotly(bfi_annual_plot)

```

**10) (3 pts) Describe the figure you have just created. First, describe the patterns. Next, suggest process-based mechanisms to describe what is happening. Why are there different patterns in the different watersheds? What could control that?**  
ANSWER:

### Correlations

Now, let's get to the correlations. We have just looked at the relationships between RR and BFI for the three watersheds, now we just need to check whether any of these relationships are significant. 

We will use the R base function `cor.test()` to calculate the correlations. We will use `method = "spearman"` within `cor.test()`. "Spearman" is a rank-based (non-parametric) correlation coefficient and does not assume a normal distribution for the data.  

The `cor.test()` does not work well with the tidyverse functions. We first need to get the data into a wide form. 

```{r}
PQ_wide <- PQ_all %>% 
  select(-Qsum, -Psum, -sum.bf, -sum.Q) %>% # remove the columns with Q, and P
  pivot_wider(names_from = "Watershed", values_from = "RR") # spread out the data for the watersheds and RRs

BUBcor <- BLANK(x = PQ_wide$BUB, y = PQ_wide$mean.bfi, method = "spearman")
BUBcor
LSCcor <- cor.test(x = BLANK, y = PQ_wide$mean.bfi, method = "spearman")
LSCcor
UTCcor <- cor.test(BLANK)
UTCcor
```

**11) (4 pts) Use the results of the corr tests and the figure of annual BFI vs annual RR to describe and interpret the results. Are there significant/insignificant correlations and which are they? In what direction are the different correlations? Last, present a plausible process-based explanation for the results you have produced.**



